On Optimality of Finite Element Integration

### 2015-11-05 TOMS_Paper/assembly.tex
Minor fixes

\begin{Def}[Perfect multilinear loop nest]
A perfect multilinear loop nest of arity $n$ is a perfect nest composed of $n$ loops, in which all of the expressions appearing in the body of the innermost loop are linear in each loop $L_i$ separately.
\end{Def}

@@

To contextualize, consider Equation~\ref{eq:quadrature} and the (abstract) loop nest implementing it illustrated in Figure~\ref{code:loopnest}. The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises a free order loop $L_e$ (over elements in the mesh), a reduction loop $L_i$ (performing numerical integration), and a multilinear loop nest $[L_j, L_k]$ (over test and trial functions). In the body of $L_k$ we have one statement evaluating the local tensor for element $e$; more statements may be present if an element is a facet (instead of a cell, as we are assuming here). An expression implements a set of monomials. The expression we are showing is in \textit{normal form}: $m$ is the (problem-dependent) number of ``terms'', $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry. Given the rigid structure of monomials (summarized in Section~\ref{sec:monomials}), we observe that any expression from any problem can be reduced to this form.

@@

We emphasize that our aim is a strategy to synthesize optimal implementations of loop nests and expressions of this particular form. In particular, we characterize optimality as follows.

@@

These transformations minimize the operation count \textit{within} the multilinear loop nest. However, when both of them are applicable, there is no simple way of predicting which one provides the largest decrease in operation count. The difference in operation count tends however to be negligible, since only the outer loops are affected.

\subsubsection{Reduction and free order loop}
\label{sec:se-rln}
Integration- and geometry-dependent expressions should logically be placed, respectively, within the reduction and free order loops. The factorization of $\sigma_{ei}$, therefore, also plays a key role for optimal sharing elimination. This is particularly important in complex forms, characterized by several coefficients, derivatives of coefficients, and extensive use of differential operators. We distinguish two cases:
\begin{itemize}
\item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the normal form reduces to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus f_i$. 
\item There is no common symbol among all terms. There still can be two terms $w_1$ and $w_2$ such that $f_i \in \sigma_{ei}^{w1}, \sigma_{ei}^{w2}$, but factorizing $f_i$ would only increase the operation count since sharing elimination opportunities across test and trial functions would be lost. On the other hand, sharing elimination must occur \textit{within} each term $\sigma_{ei}^{w}$: factorization of (derivatives of) coefficients promotes reduction-loop invariant code motion. 
\end{itemize}

>>>>

\begin{Def}[Multilinear loop nest]
A multilinear loop nest of arity $n$ is a perfect nest composed of $n$ loops, in which all of the expressions appearing in the body of the innermost loop are linear in each loop $L_i$ separately.
\end{Def}

@@

To contextualize, consider Equation~\ref{eq:quadrature} and the (abstract) loop nest implementing it illustrated in Figure~\ref{code:loopnest}. The imperfect nest $\Lambda=[L_e, L_i, L_j, L_k]$ comprises a free order loop $L_e$ (over elements in the mesh), a reduction loop $L_i$ (performing numerical integration), and a multilinear loop nest $[L_j, L_k]$ (over test and trial functions). In the body of $L_k$, one or more statements evaluate the local tensor for the element $e$. An expression implements a set of monomials. The expression shown is in \textit{normal form}: $m$ is the (problem-dependent) number of ``terms'', $\alpha_{eij}$ ($\beta_{eik}$) represents the product of the inverse Jacobian matrix for the change of coordinates with test (trial) functions, and $\sigma_{ei}$ is a function of coefficients and geometry. Given the rigid structure of monomials (summarized in Section~\ref{sec:monomials}), we observe that expressions from any problem can be reduced to this form.

@@

We recall that our aim is a strategy to synthesize optimal implementations of loop nests and expressions of this particular form. In particular, we characterize optimality as follows.

@@

These transformations minimize the operation count \textit{within} the multilinear loop nest. When both of them are applicable, there is no simple way of predicting which one provides the largest decrease in operation count. The difference in operation count, however, tends to be negligible, since only the outer loops are affected.

\subsubsection{Reduction and free order loops}
\label{sec:se-rln}
Integration- and geometry-dependent expressions should logically be placed, respectively, within the reduction and free order loops. The factorization of $\sigma_{ei}$, therefore, also plays a key role for optimal sharing elimination. This is particularly important in complex forms, characterized by several coefficients, derivatives of coefficients, and extensive use of differential operators. We distinguish two cases:
\begin{itemize}
\item For all terms $w$, there exists a common symbol $f_i \in \sigma_{ei}$ (e.g., a scalar coefficient) such that the normal form reduces to $f_i \sum_{w=1}^m \alpha_{eij}^{w} \beta_{eik}^{w} \eta_{ei}^{w}$ and $\eta_{ei} = \sigma_{ei} \setminus f_i$. 
\item There is no common symbol among all terms. There still can be two terms $w_1$ and $w_2$ such that $f_i \in \sigma_{ei}^{w1}, \sigma_{ei}^{w2}$, but factorizing $f_i$ would only increase the operation count since sharing elimination opportunities across test and trial functions would be lost. On the other hand, sharing elimination must occur \textit{within} each term $\sigma_{ei}^{w}$: factorization of (derivatives of) coefficients promotes reduction-loop invariant code motion. 
\end{itemize}
